Towards cross-middleware authentication and single sign-on for ARC Grid middleware

### 2009-02-01 doc/sec/papers/ISC09/ISC09.tex
more typesetting

\begin{figure}
\includegraphics[width=0.9\columnwidth]{HED.png}
\caption{The example of Host Environment Daemon deployed with A-REX and File services}
\label{fig:HED}
\end{figure}
HED contains a framework for implementing and enforcing authentication and authorization. Each MCC or Service has a common interface for implementing various authentication and authorization functionality. Each functionality can be implemented as a pluggable and configurable component (plug-ins) called SecHandler, which is implemented as a C++ class and provides a method for processing message that travels through MCC or Service. Each MCC or Service is usually configured with two queues of SecHandler – one for the incoming message and one for the outgoing message respectively. All SecHandler components attached to the queue are executed sequentially. If any of them fails, message processing fails as well. In Figure \ref{fig:HED}, the "AuthZ" and "AuthN" sub-modules inside MCCs and Services are examples of SecHandler.

@@

\section{Cross-Middleware Authentication and Single Sign-On}
\label{sec:siglesignon}
We investigated the authentication and single sign-on issue in the scenario when a user or user’s representative (acting on behalf of this user) needs to access resources across different middlewares. Our work focuses on the following five major aspects:
\begin{itemize}
	\item Integration of community authentication -- to utilize the standard-compatible community authentication mechanism from the Shibboleth~\cite{Shiblink} Identity Provider software when users authenticate against the Grid services.
	\item Short lived credentials service – to issue the short lived X.509 certificates for users authenticated using the community authentication, in order to interoperate with the services that depend on PKI based authentication.
	\item Credential delegation service – to process the X.509 credential delegation as a normal Web Service, in order to provide single sign-on ability.
	\item Message level authentication – to provide message level authentication by implementing WS-Security.
	\item SAML token based delegation service – to process the delegation of an WS Security SAML token in order to provide single sign-on for SOAP message level authentication.
\end{itemize}

@@

The Shibboleth provides cross-domain single sign-on and attribute-based authorization while preserving user privacy. It is based on the OASIS Security Assertion Markup Language (SAML), specifically, the new version of Shibboleth supports SAML 2.0 specification. For the authentication, the Shibboleth implements SAML2.0 web browser SSO profile, which  defines two functional components, an Identity Provider and a Service Provider. The Identity Provider (IdP) is responsible for creating, maintaining, and managing user identity, while the Service Provider (SP) is responsible for controlling access to services and resources by analyzing the SAML assertions produced and issued by the IdP upon request. In order to discover which home community does a user come from, Shibboleth specifies an optional third component called “Where Are You From?” (WAYF) service to aid in the process of IdP discovery. This IdP discovery process is also standardized and defined in SAML 2.0 specification as “Identity Provider Discovery Profile”.

@@

\begin{figure}
\includegraphics[width=0.9\columnwidth]{SAML2SSO.png}
\caption{SAML2.0 SSO profile in ARC}
\label{fig:SAML2SSO}
\end{figure}
Here is the explanation of steps shown in Figure \ref{fig:SAML2SSO}:
\begin{enumerate}
\item The client uses the user agent interface to launch a HTTP request including the IdP name (to which the user belongs) to the service side. The endpoint of the SP (Service Provider) service is the same as that of the other target services, except the last part of the endpoint is “saml2sp” which is specific for pointing to the SP Service. Note that we use Identity Provider (IdP) name here to simplify the IdP discovery process in order to avoid the IdP discovery process, because we suppose that the user who would access the target services should better know where is he from initially.
\item The SP Service searches the metadata (we use the same metadata format as defined in the Shibboleth) and gets the location of the single sign-on service (hosted in IdP) and also the location of assertion consuming service (hosted in this SP itself). Then SP Service uses obtained information to compose and issue the SAML authentication request also by utilizing own X.509 certificate (note that in the SAML2.0 SSO profile, X.509 certificates are still needed for IdP and SP) and sends the response back to the user agent.
\item User agent sends the SAML authentication request to the Identity Provider.
\item Identity Provider requires an act of authentication. The authentication mechanism is outside of the SAML2.0 SSO profile. The Shibboleth IdP implementation uses login handlers for authentication. The current user agent implementation is compatible with the Username/Password login handler of Shibboleth IdP. Through the HTTP protocol, the user agent will feed IdP with the username/password which has been given by the caller of user agent interface.
\item Once the authentication has been succeeded, the IdP issues a SAML response including SAML assertion encrypted by destination SP’s public key, and then this SAML response will be delivered by the user agent to the Service Provider.
\item The SP Service verifies and checks the SAML response, decrypts and stores the SAML assertion into session/connection context. That assertion includes the AuthnStatement and AttributeStatement elements.
\item The WS client launches the Grid/Web Service request via the same connection as the one which is used by user agent to contact SP Service.
\item The Grid/Web Service checks the AuthnStatement stored in the session context to see if the session is still valid. If valid, service handles the  request and returns the response to the WS client. Note that in the current implementation the service requires WS client to reuse same TCP connection as the one used by user agent is step 5, in order to guarantee  validity of the SSO result.
\end{enumerate}
The SP service and other functional service(s) are hosted by the same container, and they use the same X.509 credential for the transport layer security. The client authentication may be switched off, so that client doesn’t need to use any X.509 credential. Only the trusted CAs of SP and IdP services need to be configured on the client side to ensure SP and IdP can authentication.

The benefit of the SAML2.0 SSO profile worth mentioning is that the Identity Provider can cache the authentication result through session management once the user agent has succeeded to authenticate. Then for a short period this authentication result is valid so that the user agent doesn’t need to feed IdP with user’s username and password the next time it authenticates against IdP. So the user (or the client on behalf of that user) can travel across multiple security domains with only providing his name and password once, which is the characteristic of single sign-on.

@@

\begin{figure}
\includegraphics[width=0.9\columnwidth]{SLCS.png}
\caption{Short lived credential service}
\label{fig:SLCS}
\end{figure}
The SLCS service is also a Web Service implemented using ARC service interface. The SLCS client is implemented as command-line interface (CLI) utility which implements the user agent and WS client functionalities. The whole process of SLCS invocation is shown in Figure \ref{fig:SLCS} (steps 1‑8). The process resembles one shown in in Figure \ref{fig:SAML2SSO}, except that steps 7 and 8 use SLCS certificate request and response.

@@

It should be noticed that the process of composing the distinguished name (DN) for the certificate is a critical issue for the SLCS service. Since the Shibboleth Identity Provider uses the eduPerson schema~\cite{eduSchemalink} for the composing the SAML Attribute we pick the relatively distinguishable attribute “eduPersonPrincipalName” for the DN. A typical example of the eduPersonPrincipalName  could be alice@example.org. In such case composed DN is for example “/O=knowarc/OU=example.org/CN=alice”.

The obvious benefit of the SLCS service is that the user can get the X.509 credentials anywhere simply by running the SLCS client command and providing her username and password, and then access the X.509 based Grid infrastructure conveniently. Thanks to the single sign-on capability of SAML2.0 SSO profile, the user doesn’t not need to provide her username and password during valid period after the first time she succeeds to authenticate against her home IdP.

@@

\begin{figure}
\includegraphics[width=0.9\columnwidth]{Delegation_protocol.png}
\caption{Credential delegation protocol}
\label{fig:Delegation_protocol}
\end{figure}
For the delegation process, the delegation WS client functionality (inside each delegation service or the initial client) which is included to access the target delegation service (d1 or d2 in Figure \ref{fig:Delegation}), should use the user’s X.509 credential or a delegated credential (on behalf of the user) for authentication and secure communication, as well as delegate one more level of the delegated credential. For the other general  service invocation process, the WS client functionality (inside each service or the initial client) which is included to access the target service (step a1 or a2 in Figure \ref{fig:Delegation}), should also use the user’s X.509 credential or a delegated credential for authentication and secure communication.

Since each WS client accesses the target service (delegation service or other general services) on behalf of the user, the trust relationship between the user’s certificates (X.509 certificate and proxy certificate) and services is required, while the trust relationship between service’s certificate and another service’s certificate is not required.

@@

If the WS client owns an X.509 certificate or proxy certificate, the process of generating an X.509 Token for a SOAP message is: this client puts the content of certificate into X.509 token as <wsse:BinarySecurityToken>, and uses the private key (corresponding to the certificate) to create XML signature for the SOAP message body.

@@

\begin{figure}
\includegraphics[width=0.9\columnwidth]{SAMLDelegationMessage.png}
\caption{SOAP message header with SAML delegation token inside}
\label{fig:SAMLDelegationMessage}
\end{figure}
The delegation process is similar to the process in the X.509 credential delegation service, with the following differences:
\begin{enumerate}
	\item Transport level communication between delegation client and service is not based on mutual authentication between the proxy certificate and the service certificate; instead, it is based on mutual authentication between the service certificates.
	\item Delegation service will not generate any certificate request; instead, the delegation client will pick the public key from the peer certificate after a successful SSL/TLS authentication. So step 2 of the delegation protocol is just some confirmation information which informs delegation client to proceed.
	\item For those multiple intermediate services through which a SOAP message flows, delegation service does not need to appear, instead, it only needs to appear on the end of the SOAP message flow.
	\item SAML Token might include one or a few <saml:Attribute> attributes which carry the attribute information related to the subject, instead of the Attribute Certificate which is used in the VOMS proxy certificate for carrying the attribute information.
\end{enumerate}
Figure \ref{fig:SAMLDelegationMessage} shows the SAML Token (SAML Assertion) inside the SOAP message is changed due to delegation (we suppose X.509 token is used for attesting entity to authenticate against issuing authority). The SOAP message on the left side is the one which B sends to the relying party (intermediate service). B is the identity of the attesting entity’s X.509 token, while A is the identity of issuing authority, and the SOAP message is signed by B’s private key which is corresponding to the X.509 certificate that is used to create the X.509 token. On the right side, C is the identity of the delegation service (also C is the identity of the target service to which B will access), and now the SOAP message is signed by C’s private key which is corresponding to the target service’s certificate.

>>>>

\begin{figure}
\includegraphics[width=0.9\columnwidth]{HED.png}
\caption{The example of Host Environment Daemon deployed with A-REX and File services}
\label{fig:HED}
\end{figure}
HED contains a framework for implementing and enforcing authentication and authorization. Each MCC or Service has a common interface for implementing various authentication and authorization functionality. Each functionality can be implemented as a pluggable and configurable component (plug-ins) called SecHandler, which is implemented as a C++ class and provides a method for processing message that travels through MCC or Service. Each MCC or Service is usually configured with two queues of SecHandler -- one for the incoming message and one for the outgoing message respectively. All SecHandler components attached to the queue are executed sequentially. If any of them fails, message processing fails as well. In Figure \ref{fig:HED}, the "AuthZ" and "AuthN" sub-modules inside MCCs and Services are examples of SecHandler.

@@

\section{Cross-Middleware Authentication and Single Sign-On}
\label{sec:siglesignon}
We investigated the authentication and single sign-on issue in the scenario when a user or user's representative (acting on behalf of this user) needs to access resources across different middlewares. Our work focuses on the following five major aspects:
\begin{itemize}
	\item Integration of community authentication -- to utilize the standard-compatible community authentication mechanism from the Shibboleth~\cite{Shiblink} Identity Provider software when users authenticate against the Grid services.
	\item Short lived credentials service -- to issue the short lived X.509 certificates for users authenticated using the community authentication, in order to interoperate with the services that depend on PKI based authentication.
	\item Credential delegation service -- to process the X.509 credential delegation as a normal Web Service, in order to provide single sign-on ability.
	\item Message level authentication -- to provide message level authentication by implementing WS-Security.
	\item SAML token based delegation service --- to process the delegation of an WS Security SAML token in order to provide single sign-on for SOAP message level authentication.
\end{itemize}

@@

The Shibboleth provides cross-domain single sign-on and attribute-based authorization while preserving user privacy. It is based on the OASIS Security Assertion Markup Language (SAML), specifically, the new version of Shibboleth supports SAML 2.0 specification. For the authentication, the Shibboleth implements SAML2.0 web browser SSO profile, which  defines two functional components, an Identity Provider and a Service Provider. The Identity Provider (IdP) is responsible for creating, maintaining, and managing user identity, while the Service Provider (SP) is responsible for controlling access to services and resources by analyzing the SAML assertions produced and issued by the IdP upon request. In order to discover which home community does a user come from, Shibboleth specifies an optional third component called ``Where Are You From?'' (WAYF) service to aid in the process of IdP discovery. This IdP discovery process is also standardized and defined in SAML 2.0 specification as “Identity Provider Discovery Profile”.

@@

\begin{figure}
\includegraphics[width=0.9\columnwidth]{SAML2SSO.png}
\caption{SAML2.0 SSO profile in ARC}
\label{fig:SAML2SSO}
\end{figure}
Here is the explanation of steps shown in Figure \ref{fig:SAML2SSO}:
\begin{enumerate}
\item The client uses the user agent interface to launch a HTTP request including the IdP name (to which the user belongs) to the service side. The endpoint of the SP (Service Provider) service is the same as that of the other target services, except the last part of the endpoint is ``saml2sp'' which is specific for pointing to the SP Service. Note that we use Identity Provider (IdP) name here to simplify the IdP discovery process in order to avoid the IdP discovery process, because we suppose that the user who would access the target services should better know where is he from initially.
\item The SP Service searches the metadata (we use the same metadata format as defined in the Shibboleth) and gets the location of the single sign-on service (hosted in IdP) and also the location of assertion consuming service (hosted in this SP itself). Then SP Service uses obtained information to compose and issue the SAML authentication request also by utilizing own X.509 certificate (note that in the SAML2.0 SSO profile, X.509 certificates are still needed for IdP and SP) and sends the response back to the user agent.
\item User agent sends the SAML authentication request to the Identity Provider.
\item Identity Provider requires an act of authentication. The authentication mechanism is outside of the SAML2.0 SSO profile. The Shibboleth IdP implementation uses login handlers for authentication. The current user agent implementation is compatible with the Username/Password login handler of Shibboleth IdP. Through the HTTP protocol, the user agent will feed IdP with the username/password which has been given by the caller of user agent interface.
\item Once the authentication has been succeeded, the IdP issues a SAML response including SAML assertion encrypted by destination SP’s public key, and then this SAML response will be delivered by the user agent to the Service Provider.
\item The SP Service verifies and checks the SAML response, decrypts and stores the SAML assertion into session/connection context. That assertion includes the AuthnStatement and AttributeStatement elements.
\item The WS client launches the Grid/Web Service request via the same connection as the one which is used by user agent to contact SP Service.
\item The Grid/Web Service checks the AuthnStatement stored in the session context to see if the session is still valid. If valid, service handles the  request and returns the response to the WS client. Note that in the current implementation the service requires WS client to reuse same TCP connection as the one used by user agent is step 5, in order to guarantee  validity of the SSO result.
\end{enumerate}
The SP service and other functional service(s) are hosted by the same container, and they use the same X.509 credential for the transport layer security. The client authentication may be switched off, so that client doesn't need to use any X.509 credential. Only the trusted CAs of SP and IdP services need to be configured on the client side to ensure SP and IdP can authentication.

The benefit of the SAML2.0 SSO profile worth mentioning is that the Identity Provider can cache the authentication result through session management once the user agent has succeeded to authenticate. Then for a short period this authentication result is valid so that the user agent doesn't need to feed IdP with user's username and password the next time it authenticates against IdP. So the user (or the client on behalf of that user) can travel across multiple security domains with only providing his name and password once, which is the characteristic of single sign-on.

@@

\begin{figure}
\includegraphics[width=0.9\columnwidth]{SLCS.png}
\caption{Short lived credential service}
\label{fig:SLCS}
\end{figure}
The SLCS service is also a Web Service implemented using ARC service interface. The SLCS client is implemented as command-line interface (CLI) utility which implements the user agent and WS client functionalities. The whole process of SLCS invocation is shown in Figure \ref{fig:SLCS} (steps 1--8). The process resembles one shown in in Figure \ref{fig:SAML2SSO}, except that steps 7 and 8 use SLCS certificate request and response.

@@

It should be noticed that the process of composing the distinguished name (DN) for the certificate is a critical issue for the SLCS service. Since the Shibboleth Identity Provider uses the eduPerson schema~\cite{eduSchemalink} for the composing the SAML Attribute we pick the relatively distinguishable attribute ``eduPersonPrincipalName'' for the DN. A typical example of the eduPersonPrincipalName  could be alice@example.org. In such case composed DN is for example\\``/O=knowarc/OU=example.org/CN=alice''.

The obvious benefit of the SLCS service is that the user can get the X.509 credentials anywhere simply by running the SLCS client command and providing her username and password, and then access the X.509 based Grid infrastructure conveniently. Thanks to the single sign-on capability of SAML2.0 SSO profile, the user doesn't not need to provide her username and password during valid period after the first time she succeeds to authenticate against her home IdP.

@@

\begin{figure}
\includegraphics[width=0.9\columnwidth]{Delegation_protocol.png}
\caption{Credential delegation protocol}
\label{fig:Delegation_protocol}
\end{figure}
For the delegation process, the delegation WS client functionality (inside each delegation service or the initial client) which is included to access the target delegation service (d1 or d2 in Figure \ref{fig:Delegation}), should use the user's X.509 credential or a delegated credential (on behalf of the user) for authentication and secure communication, as well as delegate one more level of the delegated credential. For the other general  service invocation process, the WS client functionality (inside each service or the initial client) which is included to access the target service (step a1 or a2 in Figure \ref{fig:Delegation}), should also use the user's X.509 credential or a delegated credential for authentication and secure communication.

Since each WS client accesses the target service (delegation service or other general services) on behalf of the user, the trust relationship between the user's certificates (X.509 certificate and proxy certificate) and services is required, while the trust relationship between service's certificate and another service's certificate is not required.

@@

If the WS client owns an X.509 certificate or proxy certificate, the process of generating an X.509 Token for a SOAP message is: this client puts the content of certificate into X.509 token as $<$wsse:BinarySecurityToken$>$, and uses the private key (corresponding to the certificate) to create XML signature for the SOAP message body.

@@

\begin{figure}
\includegraphics[width=0.9\columnwidth]{SAMLDelegationMessage.png}
\caption{SOAP message header with SAML delegation token inside}
\label{fig:SAMLDelegationMessage}
\end{figure}
The delegation process is similar to the process in the X.509 credential delegation service, with the following differences:
\begin{enumerate}
	\item Transport level communication between delegation client and service is not based on mutual authentication between the proxy certificate and the service certificate; instead, it is based on mutual authentication between the service certificates.
	\item Delegation service will not generate any certificate request; instead, the delegation client will pick the public key from the peer certificate after a successful SSL/TLS authentication. So step 2 of the delegation protocol is just some confirmation information which informs delegation client to proceed.
	\item For those multiple intermediate services through which a SOAP message flows, delegation service does not need to appear, instead, it only needs to appear on the end of the SOAP message flow.
	\item SAML Token might include one or a few $<$saml:Attribute$>$ attributes which carry the attribute information related to the subject, instead of the Attribute Certificate which is used in the VOMS proxy certificate for carrying the attribute information.
\end{enumerate}
Figure \ref{fig:SAMLDelegationMessage} shows the SAML Token (SAML Assertion) inside the SOAP message is changed due to delegation (we suppose X.509 token is used for attesting entity to authenticate against issuing authority). The SOAP message on the left side is the one which B sends to the relying party (intermediate service). B is the identity of the attesting entity's X.509 token, while A is the identity of issuing authority, and the SOAP message is signed by B's private key which is corresponding to the X.509 certificate that is used to create the X.509 token. On the right side, C is the identity of the delegation service (also C is the identity of the target service to which B will access), and now the SOAP message is signed by C's private key which is corresponding to the target service's certificate.
